---
title: 'Mini-Project: COVID-19 Vaccination Rates'
author: "Vivian Cai"
date: "3/5/2022"
output: 
  pdf_document:
    latex_engine: xelatex
---

# Overview of the data

```{r}
# Import vaccination data
vax <- read.csv( "covid19vaccinesbyzipcode_test.csv" )
head(vax)
```

### Q1. What column details the total number of people fully vaccinated? 
The **persons_fully_vaccinated** column.

### Q2. What column details the Zip code tabulation area? 
The **zip_code_tabulation_area** column.

### Q3. What is the earliest date in this dataset? 
**2021-01-05**

### Q4. What is the latest date in this dataset? 
**2022-03-01** I got red mark on the course website for this question but I think the data might have updated to include more recent dates.

```{r}
tail(vax)
```
```{r}
# install.packages("skimr")
library(skimr)
skimr::skim(vax)
```

### Q5. How many numeric columns are in this dataset? 
**9**

### Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 
**18174**

### Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 
**17.17%**

```{r}
n = 18174/105840
n
```

### Q8. [Optional]: Why might this data be missing?
Some zip codes might not have residents or they failed to provide the data.

# Working with dates

```{r}
# install.packages("lubridate")
library(lubridate)
```
```{r}
today()
```
```{r}
# converting data into a lubridate format
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

```{r}
# How many days have passed since the first vaccination reported in this dataset?
today() - vax$as_of_date[1]
```
```{r}
# how many days the dataset span?
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

### Q9. How many days have passed since the last update of the dataset? 
**6**

### Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? **61** as shown by the code chunk below
```{r}
length(unique(vax$as_of_date))
```

# Working with ZIP codes

```{r}
# install.packages("zipcodeR")
library(zipcodeR)
```

```{r}
# find the centroid of the La Jolla 92037
geocode_zip('92037')
```

```{r}
# Calculate the distance between the centroids of any two ZIP codes in miles
zip_distance('92037','92109')
```

```{r}
# pull census data about ZIP code areas 
reverse_zipcode(c('92037', "92109") )
# Pull data for all ZIP codes in the dataset
# zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

# Focus on the San Diego area

```{r}
# using base R
sd <- vax[ which(vax$county == "San Diego") , ]
nrow(sd)
```


```{r}
# Using dplyr
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```
```{r}
# Using dplyr is often more convenient when we are subsetting across multiple criteria
# for example all San Diego county areas with a population of over 10,000.
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
head(sd.10)
```

### Q11. How many distinct zip codes are listed for San Diego County? **107**

```{r}
length(unique(sd$zip_code_tabulation_area))
```

### Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? **92154**

```{r}
ardsd <- sd %>%
  arrange(desc(age12_plus_population))
head(ardsd)
```
```{r}
# Using dplyr select all San Diego “county” entries on “as_of_date” “2022-02-22”
sd0222 <- sd %>%
  filter(as_of_date == "2022-02-22")

head(sd0222)
```

### Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-02-22”? **0.7041551**
```{r}
mean(sd0222$percent_of_population_fully_vaccinated, na.rm = TRUE)
```

### Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-02-22”? 
I couldn't get the same graph as shown on the instruction page. Perhaps the data was updated by the time I downloaded it.
```{r}
library(ggplot2)
ggplot(sd0222, aes(x = percent_of_population_fully_vaccinated)) +
  geom_histogram(bins = 10)
```

# Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

### Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
plt <- ggplot(ucsd) +
  aes(x = as_of_date,
      y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x = "Date", y="Percent Vaccinated", title = "Vaccination Rate of La Jolla CA 92037")
plt
```

# Comparing to similar sized areas
```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-02-22")

head(vax.36)
```

### Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
mean.36 <- mean(vax.36$percent_of_population_fully_vaccinated)
mean.36
```
```{r}
plt + geom_hline(yintercept = mean.36, color = "red", linetype = "dashed")
```

### Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”?

As shown by the output of the chunk below.

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

### Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36, aes(x = percent_of_population_fully_vaccinated)) +
  geom_histogram()
```

### Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?
The mean calculated before was 0.7334, as shown below, the 92109 and 92040 ZIP code areas are both **below** 0.7334.

```{r}
vax %>% filter(as_of_date == "2022-02-22") %>%  
  filter(zip_code_tabulation_area %in% c("92040","92109")) %>%
  select(percent_of_population_fully_vaccinated)
```

### Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)

ggplot(vax.36.all) +
       aes(
         as_of_date,
         percent_of_population_fully_vaccinated,
         group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="purple") +
  ylim(0, 1) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rates across California",
       subtitle="Only areas with a population above 36k are shown") +
  geom_hline(yintercept = mean.36, linetype="dashed")
```



### Q21. How do you feel about traveling for Spring Break and meeting for in-person class afterwards?

Based on the figures, the majority of CA has a vaccination rate above 70%. I feel relatively safe about going back in person with the vaccination status of the state. 
