---
title: "R Functions Lab (Week 5)"
author: "Vivian Cai"
date: "2/6/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Writing a Function to Grade Students' Homework

##1) Start with simple vectors
```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90) 
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80) 
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```
**Q1. Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: <https://tinyurl.com/gradeinput> [3pts]**

```{r}
# identify the lowest score first
min(student1)
min(student2)
min(student3) # min identifies NA as the lowest, that saves work for me!
```
```{r}
# finding the location of the smallest value
which.min(student1)
which.min(student2)
which.min(student3)
# which.mean doesn't identify NA as the lowest, bummer
```

```{r}
# exploring base function mean and na.rm argument
mean(student1)
mean(student2)
mean(student2, na.rm = TRUE)
# na.rm works but is not going to be a fair way for us to calculate grades, since any student with more than one missed assignment will get a higher average than what they deserve
```

```{r}
# exploring is.na()
is.na(student1)
is.na(student2)
# We can use the boolean to select and swap out all the NAs
student2[is.na(student2)]
student2[is.na(student2)] <- 0
student2
student3[is.na(student3)] <- 0
student3
# looking pretty good
```

```{r}
# now to drop the lowest scores, going back to which.min
which.min(student3) 
# now that I have NAs changed to 0, I can find the lowest score with no problem!
```

```{r}
# Time to write the function!
grade <- function(x) {
  x[is.na(x)] <- 0
  mean(x[-which.min(x)])
}

# reload example vectors
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90) 
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80) 
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)

# try out the function
grade(student1)
grade(student2)
grade(student3)
```
 Great! That was fun!
 
"Your final function should be adquately explained with **code comments** and be able to work on an example class gradebook such as this one in CSV format: <https://tinyurl.com/gradeinput>"

```{r}
# current function
#' Mean Value for Vector (Dropping the lowest value)
#' Calculates the average score after exluding the lowest score. The missing/NA values will be treated as Os.
#'
#' @param x Numeric vector of a set of values/scores
#'
#' @return Average value/score
#' @export
#'
#' @examples
#' student <- c(25, NA, NA, 60, 80, 95)
#‘ grade(student)

grade <- function(x) {
  # swapping out missing homework values to 0
  x[is.na(x)] <- 0
  # get the mean after dropping the lowest score
  mean(x[-which.min(x)])
}
```

Time to try out the gradebook!
```{r}
# reading out the csv file from a link and store it as the gradebook
url <- "https://tinyurl.com/gradeinput" 
gradebook <- read.csv(url)
gradebook
# changing row name column
gradebook <- read.csv(url, row.names = 1)
gradebook
```

```{r}
# use the apply() function on the gradebook array
apply(gradebook, 1, grade) # here 1 is for row since we want to calculated across the rows, if we want to do this across the columns, we need to put 2 for the second argument
```

**Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook? [3pts]**
```{r}
mean.score <- apply(gradebook, 1, grade)
# two ways!
which.max(mean.score)
sort(mean.score, decreasing = TRUE)
```

**Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]**

```{r}
# calculate summary stats of the gradebook
head(gradebook) # inspect first
apply(gradebook, 2, mean) # not quite
ave <- apply(gradebook, 2, mean, na.rm = TRUE)
which.min(ave)
```
```{r}
# let's try sum
apply(gradebook, 2, sum) # not quite
sum <- apply(gradebook, 2, sum, na.rm = TRUE)
which.min(sum)
```
```{r} 
# median should also work
apply(gradebook, 2, median, na.rm = TRUE)
med <- apply(gradebook, 2, median,na.rm = TRUE)
which.min(sum)
```

```{r}
# plotting things is always good
boxplot(gradebook)
```
**Q4. Optional Extension: From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)? [1pt]**
```{r}
# ?cor
head(mean.score)
gradebook[is.na(gradebook)]<- 0
cor(mean.score, gradebook$hw1)
cor(mean.score, gradebook$hw5)
```

```{r}
apply(gradebook, 2, cor, mean.score)
```

**Q5. Make sure you save your Rmarkdown document and can click the “Knit” button to generate a PDF foramt report without errors. Finally, submit your PDF to gradescope. [1pt]**


